\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-03-20 paper/main.tex
.

\subsection{Protein-centric evaluation}
The goal of \name is to find candidate drugs that target a specific
protein; however, our evaluation does not evaluate this application
but rather how \name would perform in finding plausible drug--target
interactions among all possible interactions. To provide a better
estimate on how \name performs for individual protein targets, we use
micro-averages between proteins and compute the \textit{MicroAUC};
micro-averages average the performance (true and false positive rates)
per protein instead of across all drug--protein pairs.

@@

% \subsection{Baseline model}
Furthermore, we hypothesize that some of the predictions made by our
model are the result of the biases in drug--target interaction
datasets. We design a ``na\"ive'' baseline model that predicts the
same list of proteins for each drugs based on the number of
known drug--target interactions for a protein. Formally, given lists
$D$ and $P$ of drugs and proteins and a set of known interactions
$\mathcal{I} := \{(d_i, p_i) \}$, we construct an interaction matrix
$M_{int}\in\{0,1\}^{|D|\times|P|}$ with
\begin{equation*}
	M_{ij} = \begin{cases}
		1 & \text{if } (d_i, p_j)\in \mathcal{I}\\
		0 & otherwise
	\end{cases}
\end{equation*}
describing for all drug--protein pairs whether there is a known
interaction or not. We now rank all proteins $p_j\in P$ descending by
their number of drug interactors by summing over the columns of
$M_{ij}$ and ranking these sums:
\begin{equation*}
	f: P \rightarrow \mathbb{N} \text{ with } f:p_j \mapsto \sum_{i=1}^{|D|}M_{ij}
\end{equation*}
Our ``na\"ive'' predictor $P_k$  predicts all drugs to interact with
the top $k$ targets
% , denoted by $Top_k(P)$ 
with respect to
% $M_{ij}$ from
the introduced ranking:
\begin{equation*}
	P_k: D\times P \rightarrow \{0,1\} \text{ with } P_k: d_i, p_j \mapsto \begin{cases}
		1 & \text{ if }p_j \in Top_{k}(P)\\
		0 & otherwise
	\end{cases}
\end{equation*}
with the only hyperparameter $k$.
The prediction $P_k$ is not dependent on the drug $d_i$ and will
predict the same ranked list of drugs for a given protein $p_j$;
consequently, this na\"ive predictor will not predict any useful
information about interactions between drugs and proteins, but allows
us to estimate the effect of potential biases in our evaluation
dataset.
% Also note the possibility for a similar predictor by calculating the top $k$ interacting drugs, respectively.

We apply this naive predictor on both the STITCH and Yamanishi
datasets, using the full dataset as well as 5-fold cross-validation
over drugs and drug--protein pairs.
Note, that this baseline predictor is not applicable for a protein split CV, as the number of interactions for each protein in the validation set is unknown. 
For each fold, we gradually increase $k$ to determine the best
performance for each fold. Using the full dataset, drug--target split,
and drug split, we obtain the following MacroAUC/AUROC score results. For STITCH database we obtain a performance of $0.76$ on the whole dataset, $0.70$ for the drug--target pairs and $0.73$ in case of the drug splitting scheme. While these results are fairly moderate, on Yamanishi dataset we yield MacroAUC scores of $0.88$, $0.84$ and $0.85$, for total dataset, DT pair split and drug split, respectively.

The na\"ive predictor shows great performance on Yamanishi dataset, with substantial gain in comparison to a random predictor on both datasets. In the following, we will utilize this nai\"ve predictor as baseline predictor correlating its performance to state of the art models and \name.

For comparison with the state of the art methods, we chose the best
performing methods for drug--target interaction prediction that were
previously evaluated on the Yamanishi benchmark dataset. These methods include DTIGEMS+
\cite{DTIGEMS2020} and DTI-CDF \cite{DTI-CDF2019}, which have showed
superior results in comparison to numerous works. Furthermore, we
added DTINet \cite{DTINet2017} as method for comparison which has been
used to develop a number of methods such as NeoDTI \cite{NeoDTI2019}
with similar methodology.

We evaluate all models on their recommended splitting scheme choice,
hyperparameters and folds in cross-validation, measuring their
respective AUROC. We further evaluate each model by performing a
protein-wise cross-validation determining Macro AUC/AUROC score and
$MicroAUC_p$. For this evaluation we allow stratification for the
training process but not for the validation and testing phase as
real world applications of these models would have to deal with possibly
imbalanced data.

>>>>

\subsection{Protein-centric evaluation}
The goal of \name is to find candidate drugs that target a specific
protein; however, our evaluation does not evaluate this application
but rather how \name would perform in finding plausible drug--target
interactions among all possible interactions. To provide a better
estimate on how \name performs for individual protein targets, we use
micro-averages between proteins and compute the \textit{MicroAUC} (see
Section \ref{sec:evaluation}); micro-averages average the performance
(true and false positive rates) per protein instead of across all
drug--protein pairs, and can therefore better estimate how \name
performs when tasked with finding a drug that targets a specific
protein.

@@

% \subsection{Baseline model}
Furthermore, we hypothesize that some of the predictions made by our
model are the result of the biases in drug--target interaction
datasets where hub proteins may have a large number of interactions,
or certain drugs interact with many proteins, and these may be easier
to predict than others. We design a ``na\"ive'' baseline model that
predicts the same list of proteins for each drug based only on the
number of known drug--target interactions for a protein. Formally,
given lists $D$ and $P$ of drugs and proteins and a set of known
interactions $\mathcal{I} := \{(d_i, p_i) \}$, we construct an
interaction matrix $M_{int}\in\{0,1\}^{|D|\times|P|}$ with
\begin{equation*}
	M_{ij} = \begin{cases}
		1 & \text{if } (d_i, p_j)\in \mathcal{I}\\
		0 & otherwise
	\end{cases}
\end{equation*}
describing for all drug--protein pairs whether there is a known
interaction or not. We now rank all proteins $p_j\in P$ descending by
their number of drug interactors by summing over the columns of
$M_{ij}$ and ranking these sums:
\begin{equation*}
	f: P \rightarrow \mathbb{N} \text{ with } f:p_j \mapsto \sum_{i=1}^{|D|}M_{ij}
\end{equation*}
Our ``na\"ive'' predictor $P_k$ predicts all drugs to interact with
the top $k$ targets with respect to the introduced ranking:
\begin{equation*}
	P_k: D\times P \rightarrow \{0,1\} \text{ with } P_k: d_i, p_j \mapsto \begin{cases}
		1 & \text{ if }p_j \in Top_{k}(P)\\
		0 & otherwise
	\end{cases}
\end{equation*}
with the only hyperparameter $k$.

The prediction $P_k$ is not dependent on the drug $d_i$ and will
predict the same ranked list of drugs for all proteins; consequently,
this na\"ive predictor will not predict any useful information about
interactions between drugs and proteins, but allows us to estimate the
effect of potential biases in our evaluation dataset.
% Also note the possibility for a similar predictor by calculating the top $k$ interacting drugs, respectively.

Our na\"ive predictor is not applicable for a protein split
cross-validation as the number of interactions for each protein in the
validation set is unknown.  We apply this naive predictor on both the
STITCH and Yamanishi datasets, using the full dataset as well as
5-fold cross-validation over drugs as well as over drug--protein
pairs.  For each fold, we gradually increase $k$ to determine the best
performance for each fold. Using the full dataset, drug--target split,
and drug split, we obtain the following MacroAUC results: for the
STITCH database, we obtain a performance of $0.76$ on the whole
dataset, $0.70$ for the drug--target pairs and $0.73$ in case of the
drug splitting scheme; on the Yamanishi dataset, we obtain MacroAUC
scores of $0.88$, $0.84$ and $0.85$, for the total dataset,
drug--target pair and drug split, respectively.  The na\"ive predictor
shows higher performance on the Yamanishi dataset than on STITCH, and
a substantial gain in comparison to an expected random predictor on
both datasets. In the following, we utilize this nai\"ve predictor as
baseline to compare its performance to state of the art models and
\name.

For comparison with the state of the art methods, we chose the best
performing methods for drug--target interaction prediction that were
previously evaluated on the Yamanishi benchmark dataset. These methods
include DTIGEMS+ \cite{DTIGEMS2020} and DTI-CDF \cite{DTI-CDF2019}
which have showed superior results in comparison to numerous
works. Furthermore, we added DTINet \cite{DTINet2017} as method for
comparison which has been used to develop a number of methods such as
NeoDTI \cite{NeoDTI2019} with similar methodology.

We evaluate all models on their recommended splitting scheme choice,
hyperparameters and folds in cross-validation, measuring their
respective AUROC. We further evaluate each model by performing a
protein-wise cross-validation determining the MacroAUC and
$MicroAUC_p$. For this evaluation, we allow stratification for the
training process but not for the validation and testing phase as real
world applications of these models would have to deal with possibly
imbalanced data.
